Bayesian multivariate skew-normal finite mixture model for analysis of infant development trajectories

### 2019-06-25 manuscript/manuscript.tex
section 3

\subsection{Conditional MSN Imputation}
To accommodate missing at random (MAR) responses, we propose a convenient imputation algorithm that can be implemented ``online'' as part of the Gibbs sampler. In Section 6, we discuss extensions to allow for non-ignorable missingness. Suppose $\mathbf{y}_i$ has $q_i\in (1,\ldots,J)$ observed values, denoted $\mathbf{y}^{obs}_i$, and $J-q_i$ intermittent missing values, denoted $\mathbf{y}^{miss}_i$. We can make use of the stochastic representation given in equation (\ref{eq:msnreg}) to impute $\mathbf{y}^{miss}_i$ from its conditional multivariate normal distribution given $(z_i,t_i,Y^{obs}_i)$:
\begin{eqnarray}
\mathbf{y}^{miss}_i|(z_i=k,t_i,\mathbf{y}^{obs}_i)&\sim& N_{J-q_i}(\boldsymbol\mu^{cond}_{ki},\boldsymbol\Sigma^{cond}_k),~\text{where}\nonumber\\
\boldsymbol\mu^{cond}_{ki}&=& \boldsymbol\mu^{miss} + \boldsymbol\Sigma_{12} \boldsymbol\Sigma_{22}^{-1}(\mathbf{y}^{obs}_i - \boldsymbol\mu^{obs})\\
\boldsymbol\Sigma^{cond}_k&=& \boldsymbol\Sigma_{11} - \boldsymbol\Sigma_{12}\boldsymbol\Sigma_{22}^{-1}\boldsymbol\Sigma_{21},~\text{where}\nonumber
\end{eqnarray}
$\boldsymbol\Sigma_k$ is partitioned into four sub-matrices $\boldsymbol\Sigma_{11},\boldsymbol\Sigma_{12},\boldsymbol\Sigma_{21}, \ \text{and} \ \boldsymbol\Sigma_{22}$ such that $\boldsymbol\Sigma_{11}$ is a $J-q_i \times J-q_i$ matrix containing the rows and columns of $\boldsymbol\Sigma_k$ corresponding to inicies of $\mathbf{y}_{i}$ where missingness occurs. Similarly, $\boldsymbol\Sigma_{12}$ is a $J-q_i \times q_i$ matrix containing the rows of $\boldsymbol\Sigma_k$ that correspond to missing indices of $\mathbf{y}_i$, but columns of $\boldsymbol\Sigma_k$ that correspond to observed indices of $\mathbf{y}_i$. The remaining partitions $\boldsymbol\Sigma_{21}, \ \text{and} \ \boldsymbol\Sigma_{22}$ are defined in the same manner. These results follow from conventional multivariate normal theory. An attractive feature of this imputation algorithm is that it avoids multiplicative run-time scaling in $m$, the number of imputations (Gelman \textit{et al.} 2013; Zhou and Reiter, 2010). Our approach also provides more opportunities to explore the missing data parameter space than does multiple imputation, since missing values are drawn at each MCMC iteration, and often in practice $n_{sim} >> m$, where $n_{sim}$ is the number of MCMC iterations. In Section 4, we conduct simulation studies to demonstrate that imputing the missing MSN responses improves inferences over complete case analysis.

@@

\paragraph{P\'olya--Gamma Data Augmentation for $z_i$} The sampler begins by updating the latent cluster indicators $z_i$ $(i=1,\ldots,n)$ from its multinomial logit full conditional. To facilitate sampling, we adopt an efficient data-augmentation approach introduced by Polson \textit{et al.} (2013), which expresses the inverse-logit function as a mixture P\'olya--Gamma densities. By using P\'olya--Gamma data augmentation for the multinomial model, we obtain a P\'olya--Gamma mixture of experts model, a computationally efficient way to model edge weights (\textbf{cite}). A random variable $w$ is said to follow a P\'olya--Gamma distribution with parameters $b > 0$ and $c \in {\rm I\!R}$ if
\begin{equation}
w \sim \text{PG}(b,c) \stackrel{d}{=} \frac{1}{2\pi^2}\sum_{s=1}^{\infty}\frac{g_s}{(s-1/2)^2 + c^2/(4\pi^2)}, \label{eq:pg1}
\end{equation}
where $g_s \stackrel{iid}{\sim} \text{Ga}(b,1)$ for $s = 1,...,\infty$ . Polson \textit{et al.} (2013) establish that for the $\text{PG}(b,c)$ density, and for $a,\eta \in {\rm I\!R}$,

>>>>

\subsection{Conditional MSN Imputation}
To accommodate missing at random (MAR) responses, we propose a convenient imputation algorithm that can be implemented ``online'' as part of the Gibbs sampler. In Section 6, we discuss extensions to allow for non-ignorable missingness. Suppose $\mathbf{y}_i$ has $q_i\in (1,\ldots,J)$ observed values, denoted $\mathbf{y}^{obs}_i$, and $J-q_i$ intermittent missing values, denoted $\mathbf{y}^{miss}_i$. We can make use of the stochastic representation given in equation (\ref{eq:msnreg}) to impute $\mathbf{y}^{miss}_i$ from its conditional multivariate normal distribution given $(z_i,t_i,Y^{obs}_i)$:
\begin{eqnarray}
\mathbf{y}^{miss}_i|(z_i=k,t_i,\mathbf{y}^{obs}_i)&\sim& N_{J-q_i}(\boldsymbol\mu^{cond}_{ki},\boldsymbol\Sigma^{cond}_k),~\text{where}\nonumber\\
\boldsymbol\mu^{cond}_{ki}&=& \boldsymbol\mu^{miss}_{ki} + \boldsymbol\Sigma_{12} \boldsymbol\Sigma_{22}^{-1}(\mathbf{y}^{obs}_i - \boldsymbol\mu^{obs}_{ki}) \label{eq:impute}\\
\boldsymbol\Sigma^{cond}_k&=& \boldsymbol\Sigma_{11} - \boldsymbol\Sigma_{12}\boldsymbol\Sigma_{22}^{-1}\boldsymbol\Sigma_{21},~\text{where}\nonumber
\end{eqnarray}
$\boldsymbol\Sigma_k$ is partitioned into four sub-matrices $\boldsymbol\Sigma_{11},\boldsymbol\Sigma_{12},\boldsymbol\Sigma_{21}, \ \text{and} \ \boldsymbol\Sigma_{22}$ such that $\boldsymbol\Sigma_{11}$ is a $J-q_i \times J-q_i$ matrix containing the rows and columns of $\boldsymbol\Sigma_k$ corresponding to inicies of $\mathbf{y}_{i}$ where missingness occurs. Similarly, $\boldsymbol\Sigma_{12}$ is a $J-q_i \times q_i$ matrix containing the rows of $\boldsymbol\Sigma_k$ that correspond to missing indices of $\mathbf{y}_i$, but columns of $\boldsymbol\Sigma_k$ that correspond to observed indices of $\mathbf{y}_i$. The remaining partitions $\boldsymbol\Sigma_{21}, \ \text{and} \ \boldsymbol\Sigma_{22}$ are defined in the same manner. These results follow from conventional multivariate normal theory. An attractive feature of this imputation algorithm is that it avoids multiplicative run-time scaling in $m$, the number of imputations (Gelman \textit{et al.} 2013; Zhou and Reiter, 2010). Our approach also provides more opportunities to explore the missing data parameter space than does multiple imputation, since missing values are drawn at each MCMC iteration, and often in practice $n_{sim} >> m$, where $n_{sim}$ is the number of MCMC iterations. In Section 4, we conduct simulation studies to demonstrate that imputing the missing MSN responses improves inferences over complete case analysis.

@@

\paragraph{\textbf{Step 1}: Conditional MSN Imputation} The sampler begins with imputing missing values $\mathbf{y}_i^{miss}$ conditional on current simulated values of $z_i = k$ and $t_i$ as well as the associated $\mathbf{y}_i^{obs}$. This is accomplished by drawing $\mathbf{y}_i^{miss}$ from $\text{N}_{J-q_i}(\boldsymbol\mu^{cond}_{ki},\boldsymbol\Sigma^{cond}_k)$ as described in \ref{eq:impute}.

\paragraph{\textbf{Step 3:}P\'olya--Gamma Data Augmentation for $z_i$} The sampler concludes by updating the latent cluster indicators $z_i$ $(i=1,\ldots,n)$ from its multinomial logit full conditional. To facilitate sampling, we adopt an efficient data-augmentation approach introduced by Polson \textit{et al.} (2013), which expresses the inverse-logit function as a mixture P\'olya--Gamma densities. By using P\'olya--Gamma data augmentation for the multinomial model, we obtain a P\'olya--Gamma mixture of experts model, a computationally efficient way to model edge weights (\textbf{cite}). A random variable $w$ is said to follow a P\'olya--Gamma distribution with parameters $b > 0$ and $c \in {\rm I\!R}$ if
\begin{equation}
w \sim \text{PG}(b,c) \stackrel{d}{=} \frac{1}{2\pi^2}\sum_{s=1}^{\infty}\frac{g_s}{(s-1/2)^2 + c^2/(4\pi^2)}, \label{eq:pg1}
\end{equation}
where $g_s \stackrel{iid}{\sim} \text{Ga}(b,1)$ for $s = 1,...,\infty$ . Polson \textit{et al.} (2013) establish that for the $\text{PG}(b,c)$ density, and for $a,\eta \in {\rm I\!R}$,
